---
title: "osha_analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(stringr)
```

##Load Data
```{r}
osha_analysis_data <- read_csv("../data/osha_inspections.csv")

glimpse(osha_analysis_data)

```

##Question 1:
Knowles’ accident was one of 3,203 that led to a death or “catastrophe” — defined by the U.S. Department of Labor as hospitalizations of three or more workers — and triggered an investigation by safety inspectors at the U.S. Occupational Safety and Health Administration during the first three and a half years of Donald Trump’s presidency, according to the agency’s enforcement data.

```{r}
#The first step is to check whether there were 3203 accidents that took place during the first three and a half years of Trump's presidency. The next step is to see if Knowes' company was listed in the 3203 accidents.

#According to ProPublica's repo, the "M" in insp_type refers to fatalities/ catastrophes 
#And "S" in safety_hlth column refers to Safety inspections

#PART 1: Total accidents
#First, count the total numbers of safety inspections ("S"), and catastrophes ("M")
catastrophe_inspections <- osha_analysis_data %>%
  filter(safety_hlth == "S", insp_type == "M")
count(catastrophe_inspections)

#PART 2: Trump's presidency
#Then, filter this to see how many took place in Trump's term in office -- first 3.5 years
#According to ProPublica's repo: Trump took office on 2017-01-20, and the 3.5-year period ended on 2020-07-20.
trump_inspections <- catastrophe_inspections %>%
  filter(open_date >= "2017-01-20",
         open_date < "2020-07-20")
count(trump_inspections)

#PART 3: KNOWLES' workplace
knowles_workplace <- trump_inspections %>%
  #filter, using str_detect, to look for a line in column "estab_name"
  #regex ignores case senstive characters, library stringr installed
  filter(str_detect(estab_name, regex("majestic marble", ignore_case = TRUE)))
knowles_workplace

```
##Question 2:
In 2019, OSHA’s safety inspectors conducted 962 investigations into fatal or catastrophic workplace incidents — the highest number since the agency began publishing the data in 2011.

```{r}
year_fatal_cat_insp <- osha_analysis_data %>%
  filter(insp_type == "M", safety_hlth == "S") %>%
  mutate(year = year(open_date)) %>%
  group_by(year) %>%
  select(year, insp_type) %>%
  count(insp_type = "M")%>%
  filter(year == "2019")

year_fatal_cat_insp
  
```

##Question 3:
The agency has conducted slightly fewer safety inspections during the first three years of Trump’s presidency than during a comparable period at the end of President Barack Obama’s second term, even though the labor force grew by 16 percent, according to a Center for Public Integrity analysis of the agency’s inspection data.


```{r}
#Step 1: calculate number of SAFETY inspections conducted in 3 years of Trump's presidency
#"S" in safety_hlth column refers to Safety inspections
trump_inspections_three_years <- osha_analysis_data %>%
  filter(safety_hlth == "S")%>%
  filter(open_date >= "2017-01-20",
         open_date < "2020-01-20",)
count(trump_inspections_three_years)

#Answer: 88354	

#Step 2: Obama 3 years
obama_inspections_three_years <- osha_analysis_data %>%
  filter(safety_hlth == "S")%>%
  filter(open_date>="2014-01-20",
         open_date < "2017-01-20")
count(obama_inspections_three_years)
#Answer: 93992


```




