---
title: "osha_analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
```

##Load Data
```{r}
osha_analysis_data <- read_csv("../data/osha_inspections.csv")

glimpse(osha_analysis_data)

```

Knowles’ accident was one of 3,203 that led to a death or “catastrophe” — defined by the U.S. Department of Labor as hospitalizations of three or more workers — and triggered an investigation by safety inspectors at the U.S. Occupational Safety and Health Administration during the first three and a half years of Donald Trump’s presidency, according to the agency’s enforcement data.

```{r}
#total_inspections_trump <- osha_analysis_data %>%
  #filter(open_date >= "2017-01-20") %>%
  #filter(open_date <= "2020-01-20")

trump_inspections <- osha_analysis_data %>%
  filter(safety_hlth == safety,
           open_date >= "2017-01-20",
         open_date < "2020-01-20",
         insp_scope != "D") # no non-inspections
count(trump_inspections)

```

