---
title: "osha_analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
```

##Load Data
```{r}
osha_analysis_data <- read_csv("data/osha_inspections.csv")

glimpse(osha_analysis_data)

```

3) The agency has conducted slightly fewer safety inspections during the first three years of Trump’s presidency than during a comparable period at the end of President Barack Obama’s second term, even though the labor force grew by 16 percent, according to a Center for Public Integrity analysis of the agency’s inspection data.

```{r}
osha_analysis_data %>%
  group_by(insp_scope, why_no_insp) %>%
  count()
```

```{r}
trump_begin_date <- "2017-01-20"
trump_end_date <- "2020-01-20"

trump_inspections <- osha_inspections %>%
  filter(safety_hlth == "S") %>%
 filter(open_date >= trump_begin_date,
        open_date < trump_end_date)
       # insp_scope != "D") # no non-inspections

count(trump_inspections)
  
#Q: How many safety inspections were conducted in the first three years of Trump's presidency?
#A:88354

#Question for Sean: What does line 49 do, becuase that's changing my outcome.
```

```{r}
obama_begin_date <- "2014-01-20"
obama_end_date <- "2017-01-20"

obama_inspections_2t <- osha_inspections %>%
  filter(safety_hlth == "S") %>%
 filter(open_date >= obama_begin_date,
        open_date < obama_end_date)
       # insp_scope != "D") # no non-inspections

count(obama_inspections_2t)

#Q: How many safety inspections were conducted in the last three years of Obama's second term?
#A:93992
```

```{r}
(count(obama_inspections_2t)) - (count(trump_inspections))

#Q: What's the difference in inspections between Trump and Obama?
#A:5638. OSHA conducted 5638 more safety inspections during the last three years of Obama's second term than in the first three years of Trump's presidency. A slightly larger amount.
```

### Sentence 3

"The agency has conducted slightly fewer safety inspections during the first three years of Trump’s presidency than during a comparable period at the end of President Barack Obama’s second term, even though the labor force grew by 16 percent, according to a Center for Public Integrity analysis of the agency’s inspection data."

```{r}
#Then, filter this to see how many took place in Trump's term in office -- first 3.5 years
#According to ProPublica's repo: Trump took office on 2017-01-20, and the 3.5-year period ended on 2020-07-20.

trump_begin_date <- "2017-01-20"
trump_end_date <- "2020-01-20"

#According to the Center for Public Integrity's repo, the "M" in insp_type refers to fatalities/ catastrophes 
#And "S" in safety_hlth column refers to Safety inspections

# Part 1: Inspections conducted during Trump's presidency
# create an object, trump_inspections
trump_inspections <- osha_analysis_data %>%
  filter(safety_hlth == "S") %>%
  #filter to see inspections conducted during Trump's presidency.
  filter(open_date >= trump_begin_date,
        open_date < trump_end_date,
        insp_scope != "D") # this means that were no inspection conducted
count(trump_inspections)
  
#Q: How many safety inspections were conducted in the first three years of Trump's presidency?
#A:88354
```

```{r}
# Part 2: Inspections conducted during Obama's presidency
obama_begin_date <- "2014-01-20"
obama_end_date <- "2017-01-20"
# create an object, obama_inspections
obama_inspections <- osha_analysis_data %>%
  filter(safety_hlth == "S") %>%
   filter(open_date >= obama_begin_date,
          open_date < obama_end_date,
          insp_scope != "D") # no non-inspections
count(obama_inspections)
#Q: How many safety inspections were conducted in the last three years of Obama's second term?
#A:93992
```

```{r}
#Part 3: Compare the two

#x = obama - trump
x <- (count(obama_inspections)) - (count(trump_inspections))

(count(trump_inspections) - count(obama_inspections)
) / count(obama_inspections) * 100
#Q: What's the difference in inspections between Trump and Obama?
#A:5638. OSHA conducted 5638 more safety inspections during the last three years of Obama's second term than in the first three years of Trump's presidency. A slightly larger amount.
```
